%
% ---------- header -----------------------------------------------------------
%
% project       kaneton
%
% license       kaneton
%
% file          /home/mycure/kaneton/view/book/assignments/future-k3/k4.tex
%
% created       julien quintard   [fri nov 28 05:25:37 2008]
% updated       julien quintard   [thu apr  9 23:25:21 2009]
%

%
% ---------- k4 ---------------------------------------------------------------
%

\chapter{k4}
\label{chapter:k4}

\name{k4} gives students the possibility to experiment through the design
and development of a major system components.

\newpage

%
% ---------- text -------------------------------------------------------------
%

%
% objectives
%

\section{Objectives}

Through this project, students are going to explore the multiple ways of
providing a major operating system, being the file system or network
abstraction.

Students will have to study the pros and cons of every technique in order to
choose the right algorithms to use in their implementation.

Therefore, in this project, students are free to design and implement whatever
they wish as long as it provides the required feature.

Indeed, students having decided to implement a network device driver for
\name{k3} will be asked to design and implement a network stack while students
having focused on a storage device will have to design and implement a file
system.

%
% requirements
%

\section{Requirements}

This stage requires a working device driver being for a storage or network
device.

The file system will rely on the disk driver for storing blocks of data while
the network stack will rely on the network device driver for sending and
receiving \name{Ethernet} frames.

%
% snapshot
%

\section{Snapshot}

For this project, nothing is provided to the students but the freedom to
innovate.

%
% assignments
%

\section{Assignments}

% design

\subsection*{Design}

Students are asked to study the different data structures and algorithms
that could be used for providing the required abstraction so that
a design is sketched.

% implementation

\subsection*{Implementation}

Students are then asked to implement the system they designed.

In order to illustrate the functionalities of their system, students
should also provide user-level applications.
