%
% ---------- header -----------------------------------------------------------
%
% project       kaneton
%
% license       kaneton
%
% file          /home/mycure/kaneton/view/book/assignments/k1.tex
%
% created       julien quintard   [tue feb  1 21:40:19 2011]
% updated       julien quintard   [mon mar  7 14:45:23 2011]
%

%
% ---------- k1 ---------------------------------------------------------------
%

\chapter{k1}
\label{chapter:k1}

\name{k1} is the first stage taking place within the \name{kaneton}
microkernel. This stage focuses on the event processing mechanism, one of
the most fundamental kernel component. This mechanism provides the kernel
the capacity to react to internal or external stimuli and to perform
computations based on their nature.

The event processing mechanism enables the kernel to catch errors, known
as \name{exceptions} on \name{Intel} architectures, such as division by
zero, page faults \etc{} but also to communicate with external devices.
Indeed, although the easiest way to communicate with devices consists in
probing \name{I/O} ports until the device's state changes to something
expected, \name{interrupt}s provide a mechanism consisting for devices
to generate asynchronously an event which will, as its name suggests, suspend
the kernel's activity to handle the interrupt.

Such events are extremely important as they are also used to simulate
multitasking \ie{} the ability to execute multiple threads of execution
in ``parallel''. Note that the mechanism underlying this concept will be
discussed in \name{k3}.

In this stage, students will therefore address the following concepts:

\begin{itemize}
  \item
    \term{Event Mechanism Theory}

    \-

    The way event processing mechanisms are implemented depending on the
    architecture will be studied through the lectures.
  \item
    \term{IDT - Interrupt Descriptor Table}

    \-

    The \name{IDT} will be manually build and managed in order to specify
    the \name{CPU} how to handle the interrupts being exceptions,
    \name{IRQ - Interrupt Request}s \etc{}
  \item
    \term{ISR - Interrupt Service Routine}

    \-

    Once an interrupt is triggered, a handler---referred to as an \name{ISR}
    on \name{Intel} architectures---is launched. This handler is responsible
    for saving and restoring the state of the interrupted execution context
    among others.
  \item
    \term{PIC - Programmable Interrupt Controller}

    \-

    The student will also discover the existence of a chipset known as the
    \name{PIC} whose purpose is directly linked to the interrupt handling.
\end{itemize}

\newpage

%
% ---------- text -------------------------------------------------------------
%

%
% submission
%

\section{Submission}

Unlike \name{k0}, students are not supposed to provide a specifically built
snapshot. Instead, the test system will archive the student's code in a proper
way and submit it.

For the process to work, students must first take care to remove the \name{k0}
snapshot which may be lying at the following location:
\location{test/client/snapshot.tar.bz2}. If this snapshot is not removed, the
test system will use it for the submission process which will result in
\name{k0}'s code being submitted for the \name{k1} stage.

Besides, students should switch their capability so as to use the one provided
for the stage \name{k1} and above.

%
% requirements
%

\section{Requirements}

Every student should read the \name{IA-32 Architectures Software
  Developer's Manual, Volume 3A: System Programming Guide, Part 1} which
is available on the \name{Wiki}. More precisely students should focus
on the chapter \name{Interrupt and Exception Handling} which contains
all the necessary material to complete this stage.

Finally, if not already the case, students should also read the chapters
\name{Protected-Mode Memory Management} and \name{Protection} as interrupt
handling makes use of the protected mode capabilities.

%
% assignments
%

\section{Assignments}

The \name{k1} stage, focusing on the event processing mechanism, takes place
entirely in the \name{event} manager. However, since every processing is
highly machine-dependent, related code can also be found in the \name{glue},
component whose main task is to redirect calls to the functionalities provided
by the \name{platform} and \name{architecture}.

Although, as students will soon notice, the provided snapshot includes a
complete event manager's core, the glue is completely empty while the
architecture does not provide any functionality related to the event
processing mechanism.

Therefore, the \name{k1} assignments consist for the student to equip the
kaneton educational implementation with the necessary functionalities making
the event manager fully operational.

Though this stage tries to give as much freedom as possible to students
regarding their implementation, skeleton source files have been provided, hence
acting as entry point for students to start experimenting.

% core
\subsection{Core}

Although the event manager is obviously the most important manager when
it comes to the event processing mechanism, the student should consider
the timer manager as well.

Indeed, the timer manager relies on the underlying hardware timer event
for providing the possibility to trigger functions after some time. The
student may be interested in this manager as being a perfect application
of the event manager.

% event
\subsubsection*{event}

\begin{itemize}
  \item
    \location{kaneton/core/event/event.c}
  \item
    \location{kaneton/core/include/event.h}
\end{itemize}

The event manager's core provides the high level interface enabling the kernel
and its servers to plug an event handler with a specific event number \ie{}
event identifier. Note that event identifiers are machine-dependent; for
instance the event \code{\#32} on \name{Intel} architectures represents the
timer. The event manager's core thus does not understand the meaning of the
event identifiers and it is therefore the responsability of the machine,
through the glue component, to perform the necessary operations according
to the underlying platform/architecture.

Although the manager's core is complete, the student will probably find it
useful to browse its source code to better understand its purpose.

% glue
\subsection{Glue}

As mentioned earlier, the timer manager relies on the hardware timer event.
The student will notice in
\location{kaneton/machine/glue/ibm-pc.ia32/educational/timer.c} that this
event is actually reserved and linked with a handler located within the
timer manager's core.

% event
\subsubsection*{event}

\begin{itemize}
  \item
    \location{kaneton/machine/glue/ibm-pc.ia32/educational/event.c}
  \item
    \location{kaneton/machine/glue/ibm-pc.ia32/educational/include/event.h}
\end{itemize}

The event manager's glue makes the binding between a core request and the
underlying platform and architecture functionalities.

The student will notice that the glue is empty when it comes to the
event manager. Indeed, it is the student's responsability to complete the
glue in order to equip some of the event manager's functionalities with a
machine-specific behaviour.

One can notice an empty array referred to as \code{glue\_event\_dispatcher}.
The role of such so called dispatchers is to receive the requests from
the core, \ie{} whenever the core calls \code{machine\_call()}, and
handles them by taking into account both the platform and the microprocessor
architecture, \name{IBM PC} and \name{IA-32} for instance.

The dispatcher must be filled with the symbols of the functions the
glue wishes to support. The dispatcher composition is located in the
core header of the manager. For instance, considering the event manager,
one can retrieve the event dispatcher composition \code{d\_event} in
the header file \location{kaneton/core/include/event.h}. For instance,
should one want to catch the calls to \code{event\_reserve()} in
the glue, a function would have to be written, say
\code{glue\_event\_reserve()}, and registered in the
\code{glue\_event\_dispatcher} array at the sixth position.

% architecture
\subsection{Architecture}

The architecture must be completed in order to provide functionalities for
controlling the microprocessor's capability to handle the interrupts and
exceptions.

For that purpose, skeleton source files are provided for the student to
implement functions regarding the management of the \name{IDT - Interrupt
Descriptor Table} and the set up of the interrupt handlers.

% idt
\subsubsection*{idt}

\begin{itemize}
  \item
    \location{kaneton/machine/architecture/ia32/educational/idt.c}
  \item
    \location{kaneton/machine/architecture/ia32/educational/include/idt.h}
\end{itemize}

The file \location{idt.c} must be completed by students in order to
implement functions related to the \name{IDT} management such as building
an \name{IDT} from a given address, inserting an entry in the \name{IDT}
according to some arguments, deleting an existing \name{IDT} entry \etc{}

Note that in order to create \name{IDT} entries, students should have a look
at the function \code{architecture\_gdt\_selector()} which enables the
caller to generate a segment selector according to several parameters.

% handler
\subsubsection*{handler}

\begin{itemize}
  \item
    \location{kaneton/machine/architecture/ia32/educational/handler.c}
  \item
    \location{kaneton/machine/architecture/ia32/educational/include/handler.h}
\end{itemize}

\name{Intel} interrupts can be classified according to their purpose into
one of the following categories: exceptions, \name{IRQ - Interrupt Request}s,
\name{IPI - Inter-Processor Interrupt}s and system calls \aka{}
\name{syscall}s. Noteworthy is that \name{IPI}s can be ignored since the
educational implementation only supports mono-processor architectures.

The objective is for students to implement the low-level interrupt handlers
which are triggered whenever the associated interrupt is received. Such a
handler, sometimes referred to as \name{lower half}, performs two basic
operations. First, the environment is secured for the execution of the
interrupted thread to be resumed properly, once the interrupt treated.
Second, the high-level event handler function registered through
\code{event\_reserve()} is triggered; this function is sometimes referred
to as the \name{higher half}.

The file \location{handler.c} should thus contain the interrupt handlers
associated with all the interrupts the kernel wishes to handle. Besides,
this file should make use of the functions provided in \location{idt.c}
for setting up the system's \name{IDT}.

% architecture
\subsubsection*{architecture}

\begin{itemize}
  \item
    \location{kaneton/machine/architecture/ia32/educational/architecture.c}
  \item
    \location{kaneton/machine/architecture/ia32/educational/include/architecture.h}
\end{itemize}

The student should also take care to record, whenever appropriate, the
error code associated with the interrupt in \code{\_architecture.error}.

A special attention must be given to the \code{\_architecture} structure.
Indeed, because this structure is being used by the testing system, its
organisation should never be modified.

% platform
\subsection{Platform}

% pic
\subsubsection*{PIC}

\begin{itemize}
  \item
    \location{kaneton/machine/platform/ibm-pc/pic.c}
  \item
    \location{kaneton/machine/platform/ibm-pc/include/pic.h}
\end{itemize}

The platform plays a minor role in the event processing mechanism. However,
students should have a look at the \name{PIC - Programmable Interrupt
Controller} which may prove useful in this stage.

%
% example
%

\section{Example}

This section presents an example which students can use to better understand
the use of the event manager. This example illustrates a breakpoint exception
being generated manually through the \code{int} assembly instruction.
The exception is then caught by the kernel which notices that an event handler
has been registered and thus triggers it.

\begin{verbatim}
void                    exception_bp(t_id                       id,
                                     t_vaddr                    data)
{
  printf("[event %qd] breakpoint exception caught\n",
         id);
}

void                    example(void)
{
  if (event_reserve(ARCHITECTURE_IDT_EXCEPTION_BP,
                    EVENT_TYPE_FUNCTION,
                    EVENT_ROUTINE(exception_bp),
                    EVENT_DATA(NULL)) != STATUS_OK)
    {
      printf("[event_reserve] error");
      return;
    }

  asm volatile("int $3");

  if (event_release(ARCHITECTURE_IDT_EXCEPTION_BP) != STATUS_OK)
    {
      printf("[event_release] error");
      return;
    }
}
\end{verbatim}

Note however that this example is specific to exceptions. In order to test
\name{IRQ}s, students should first activate the event system by invoking the
\code{event\_enable()} function.

%
% advices
%

\section{Advices}

This section contains advices that students are welcome to consider:

\begin{itemize}
  \item
    Students should write macro-functions for setting the fields within
    the \name{IDT} entries, making it easier to transform a single address
    into scattered fragments placed at different position within an entry.

    \-

    One can take the example of \name{GDT}-specific macro-functions
    which are located in \\
    \location{kaneton/machine/architecture/ia32/educational/include/gdt.h}

    \-

    In addition, the opposite should also be set up in order to easily
    retrieve a field from an \name{IDT} entry.
  \item
    The interrupt handlers located in \location{handler.c} should be
    composed of two parts.

    \-

    The first one, composing the entry point of the \name{ISR - Interrupt
    Service Routine}, should be written in assembly in order to prevent
    the compiler from generating additional instructions.

    \-

    The second one, called by the first one and written in \name{C}, triggers
    the high-level event handlers registered for this event.
  \item
    The assembly part of the interrupt handlers are actually all very similar,
    differing slightly depending on the nature of the interrupt: exception,
    \name{IRQ}, \name{IPI} \etc{} and depending on the presence of an
    error code.

    \-

    Therefore, instead of writing all these assembly functions manually,
    one could rely on a macro-function for generating the assembly
    code automatically.
  \item
    In order to ease the debugging process, students should get used to
    writing functions for dumping the state of the processor's structure,
    such as the \name{IDT}.
  \item
    An interrupt handler, noticing that no high-level handler has been
    registered for this event identifier, should display a message on the
    console warning that this interrupt was unexpected.

    \-

    This way, the developer will easily notice exceptions occuring as well
    as unhandled IRQs.
  \item
    The \code{event\_enable()} function must behave in a particular way
    for the test system to work properly. In addition to activating the
    interrupt mechanism, this function must also trigger an event
    right away.

    \-

    Indeed, most tests call \code{event\_enable()} followed by a call
    to \code{event\_disable()}, assuming that at least one timer interrupt
    will have been triggered in between for the actual test to be
    launched.

    \-

    Thus, if it happens that no timer interrupt is generated between
    the two calls, the test would actually fail. Students are therefore
    advised to force the generation of such an event.
  \item
    Students must understand the differences between exceptions and
    interrupts, especially in the context of kaneton.

    \-

    For instance, although exceptions can be triggered at any time,
    interrupts require the event mechanism to be enabled, through
    \code{event\_enable()}, not to mention that interrupts must
    not be masked through the \name{PIC}.
\end{itemize}
