%
% ---------- header -----------------------------------------------------------
%
% project       kaneton
%
% license       kaneton
%
% file          /home/mycure/kaneton/view/book/assignments/future/k3.tex
%
% created       julien quintard   [fri nov 28 05:25:37 2008]
% updated       julien quintard   [fri nov 28 05:35:43 2008]
%

%
% ---------- k3 ---------------------------------------------------------------
%

\chapter{k3}
\label{chapter:k3}

\name{k3} consists for students to study drivers development.

\newpage

%
% ---------- text -------------------------------------------------------------
%

%
% objectives
%

\section{Objectives}

The main objectives of this project is to understand who drivers are
generaly developed inside micro-kernels.

Students will be asked to implement their own driver
of the device of their choise. It could be a ATA driver, ehternet
driver, etc, \ldots{}. The device must support DMA transfert and the
driver \textbf{must} support this capability.

The driver will run under a less privileged level than the core,
therefore students will have to implement a communication mechanism 
on the top of the existing IPC infrastructure.

In this project, students are free to design and implement whatever
they wish.

%
% requirements
%

\section{Requirements}

This stage requires a working core (IPC, Memory, Scheduler).

%
% snapshot
%

\section{Snapshot}

For this project, nothing is provided to the students but the freedom to
innovate.

%
% assignments
%

\section{Assignments}

% design

\subsection*{Design}

Students are asked to follow the micro-kernel design for this driver.
Their futur driver have to run under a less privileged level than the
core.

% implementation

\subsection*{Implementation}

Students are then asked to implement a driver for the device of their
choice. The driver \textbf{must} support DMA.

If needed, in order to support the driver, extra componnants could be
integrated inside the core. If so students will be asked to argue about the
reason why those componnants have to be inside the core.

In order to illustrate the functionalities of their driver, students
should also provide user-level applications.
