%
% ---------- header -----------------------------------------------------------
%
% project       kaneton
%
% license       kaneton
%
% file          /home/mycure/kaneton/view/book/development/community.tex
%
% created       julien quintard   [sun may 20 18:08:17 2007]
% updated       julien quintard   [fri aug  1 15:51:12 2008]
%

%
% ---------- community --------------------------------------------------------
%

\chapter{Community}
\label{chapter:community}

This chapter discusses what is a community and how contributors must integrate
the kaneton community.

\newpage

%
% ---------- text -------------------------------------------------------------
%

kaneton can obviously be considered as an open source community although the
produced soure code is actually not open source.

Driving an open source community is complicated since people have different
personal goals at working on a free project. Some people contribute for
the knowledge, other for building the next generation system, other to
provide free open source softwares, other to become famous \etc{}

kaneton is a community driven microkernel that acts with the best interest of
the students at heart. Rules and regulations that keep the project moving
forward are fundamental even if the size of the kaneton community is
relatively small, for now.

Indeed, the main objective of the kaneton project remains to be as
understandable as possible in order to lead students to implement parts of
it very quickly.

The remaining of this chapter draws a list of rules contributors must agree
to respect.

% objective

\subsubsection{Objective}

kaneton aims at providing a powerful, understandable and maintainable
microkernel. This objective must be kept in mind of every contributor
since many design and implementation were/are/will be made according to
this precise objective.

Note that the kaneton microkernel does not intend to be a desktop operating
system nor an as optimised as the Linux operating system. Every contributor
should be well-aware of that in order to avoid behaviours stating that a
feature is fundamental or useless for performance concerns, for instance.

This rule does not prohibit people to suggest ideas but instead regulates
behaviours of people who wants to change major design and/or implementation
choices for bad reasons.

% behaviour

\subsubsection{Behaviour}

Open source projects does not mean constraint-free projects. The kaneton
people, whilst being relatively young, try to act for the project's good
by behaving remarkably in the kaneton community.

Therefore, contributors are asked to do the same by avoiding some bad/young
behaviours.

\begin{enumerate}
  \item
    Follow the rules. People who do not respect these rules could be banned
    from the kaneton project.
  \item
    Avoid the \textit{cowboy} behaviour consisting for a contributor to
    implement a feature without discussing about its usefulness with the
    community first. Another effect of this behaviour can be to distract
    the contributors from its major focuses.
  \item
    Always act and think in the project interest rather than your personal
    interest.
  \item
    Respect the other kaneton people, especially the ones who have worked
    on this project for a long time and who made this whole project possible.
    When people disagree, they are asked to do it respectfully.
  \item
    Take your responsability when you realise that you did something wrong:
    insults, mistakes in an implemented feature \etc{}
  \item
    \textit{``The Perfect is the Enemy of the Good!''}: even nice
    contributors can unintentionally do bad things by being perfectionists
    and/or too much into the project and/or obsessed with process.
  \item
    ... \textit{Politness}, \textit{Respect}, \textit{Trust} and
    \textit{Humility} are the key qualities that make a good contributor in
    any community.
\end{enumerate}

% communication

\subsubsection{Communication}

The communication mainly takes two forms in the kaneton microkernel project:
the \name{mailing-list} for internal communication and the \name{kaneton
website} for external communication. The \name{Developers Intranet} is another
source of communication as well as the commit logs \etc{}

The rules related to these tools are described in \reference{Chapter
\ref{chapter:tools}} and will therefore not be discussed here.

Every contributor must take the time to communicate in the mailing-list as well
as through the public website. Indeed, kaneton people must, frequently,
briefly describe what they are working on in order to inform the other
contributors who are not aware of everyone's current work. Note that these
kind of messages are very different from messages generated by repository
commits. Indeed, while these commit messages indicate a modification, they
do not describe the whole work behind them.

Additionally, contributors can communicate informing the kaneton community of
their unavailability for the next two months, for instance. This behaviour
allows people to be aware that some tasks will not be done because the
contributors in charge of it cannot work at this moment.

Althoug people are highly welcomed to communicate, some rules apply to
avoid further problems.

First, any new contributor should obviously read the kaneton documents before
asking anything which has already been discussed and decided, unless the
developer knows exactly what he/she is talking about. Indeed, asking too many
questions about the source code is a form of disrespect to the other
contributors. Moreover, many things can be found out just looking at the
kaneton documents and/or source code.

Although people are asked to communicate, people are also asked to act
respectully. Contributors should not respond to every message in every
discussion, this is a ridiculous behaviour. Instead, every developer should
carefully read the discussion, think about its response and then write a clear
message stating his point of view, ideas \etc{}

Depending on the contributor status, reading the mailing-list frequently is
absolutely fundamental as some people rely on other contributors' decisions,
advices \etc{}

Finally, the mailing-list must be considered as the official internal
communication medium. If contributors previously had a private conversation,
in real-life or on \name{IRC} for instance, the discussion must be reported
on the mailing-list so that everyone can take these new ideas into account.

% work

\subsubsection{Work}

Working on the kaneton microkernel project does not imply low-level programming
all the time. Indeed, the kaneton project is composed of two parts: the
kaneton microkernel research project and the kaneton educational project.

Although the microkernel research project requires highly skilled programmers,
it also needs documentations and some tools for performing important tasks as
diverse as generating the prototypes, testing the microkernel behaviour,
generating the documentation \etc{}

The educational project essentially needs documentation, lecture materials
and tools for managing the project: testing the students' implementation,
checking if some students cheated and many others.

This means that kaneton people must contribute to every type of task
that need to be done. Also, contributors are asked to well document
any work they have done including source code comments but also through
kaneton official documents which are then made available on the website.

% supervisor

\subsubsection{Supervisor}

A supervisor is attached to each new contributor for a certain, not fixed,
period of time. The role of this supervisor is to advise, correct and
encourage the newcomer so that it integrates well the kaneton community.

The supervisor will be someone having a well understanding of the new
contributor's project. A high level of communication is expected between
the supervisor and the contributor. A direct phone communication is highly
recommanded through \name{Skype} for instance.

% trust

\subsubsection{Trust}

A new contributor joining the kaneton project must acquire the trust of
the community. Therefore, the contributor first does not get any access
to restricted tools and must submit patches to its supervisor who review
them, taking care of advising the newcomer of its mistakes.

At the end of the test period, the community decides whether the contributor
is helpful to the project or not. Then, the contributor is either granted
of full access to kaneton tools or eliminated from the project.
