%
% ---------- header -----------------------------------------------------------
%
% project       kaneton
%
% license       kaneton
%
% file          /home/mycure/kaneton/view/book/development/environment.tex
%
% created       julien quintard   [sun may 20 14:49:26 2007]
% updated       julien quintard   [mon may  4 19:41:14 2009]
%

%
% ---------- environment ------------------------------------------------------
%

\subsection{Environment}
\label{section:environment}

Over the years, the kaneton microkernel evolved, starting with a very simple
introduction to low-level programming and finally to a complete microkernel
development.

kaneton people wanted to lead students to a complete microkernel development
to finally introduce distributed computing. This would not have been possible
if students had to build an entire development environment because developing
such an environment is a whole project by itself.

As a result, kaneton people decided to provide students a complete development
environment. The kaneton development environment is composed of make files,
python scripts and configuration files. This development environment can be
considered as one of the major kaneton tools since contributors use it
everytime.

The kaneton development environment aims at providing an easy and portable
way for managing the kaneton microkernel project from a development point
of view. Therefore, the kaneton environment provides everything necessary
for compiling, assembling, \etc{} These tasks highly rely on the underlying
running operating system as well as on the kaneton microkernel's target
microprocessor. Moreover, the user could need to redefine some behaviours
depending on its personal operating system configuration to use a specific
C compiler for instance.

The kaneton development environment provides a layered organisation of
profiles, each profile defining variables and functions used by the final
environment engine. The goal of the layered model is to allow layers to
override the definitions of lower-level layers.

%
% profiles
%

\subsubsection{Profiles}

The development environment is composed of profiles including a \name{host}
profile which describes the behaviour of the underlying operating system, a
\name{kaneton} profile which parameters the kaneton microkernel and a
\name{user} profile which permits the user to redefine lower-level layers'
declarations.

These profiles eventually hold sub-profiles which define variables and
functions. These actual profiles are accessed according to user-defined
shell variables.

% host

\subsubsubsection{Host}

The \name{host} profile essentially describes how to perform basic tasks:
compile, assemble, change the current directory, display a message \etc{} These
tasks rely on the operating system currently running as well as on the target
processor which kaneton will be built for. For these reasons, there are
several host sub-profiles.

Let us suppose a developer is running a \name{Linux} operating system and
that kaneton will be built for running on a \name{PowerPC} microprocessor. In
such a case, the C compiler program will be different depending on the
microprocessor \name{Linux} is running on. Indeed, if Linux is running on
a \name{PowerPC} microprocessor, then using the default compiler should
produce \name{PowerPC} object files. This is well-known to be the common
compiling way. On the other hand, if \name{Linux} is running on a
different microprocessor, then a cross-compiler must be used to produce
binary objects targeting a specific different microprocessor architecture,
the \name{PowerPC} architecture in our example.

To avoid this issue, a \name{host} sub-profile name is composed of two parts
separated by a slash. The first part is the name of the operating system and
the latter is a pair source/target processors separated by a period. For
example, \name{linux/ia32.ppc} names a \name{host} profile representing a
\name{Linux} operating system on a \name{Intel 32-bit} microprocessor
which aims at building a kaneton microkernel for a \name{PowerPC}
target architecture. Needless to say that \name{linux/ia32.ia32} represents
a non cross-compiling environment.

To avoid configuration duplications, it is common to see the configuration
file of a host sub-profile to include files of the parent directory as
shown below:

\begin{verbatim}
  linux/
    linux.desc
    linux.conf
    linux.mk
    linux.py
    ia32.ia32/
      educational -> .
      optimised -> .
      smp -> .
      ia32.desc
      ia32.conf
      ia32.mk
      ia32.py
    ia32.mips64/
      mips64.desc
      mips64.conf
      mips64.mk
      mips64.py
\end{verbatim}

Note that the files \location{linux.*} are not directly included by the
development environment engine since \name{linux} is not a valid host
profile name.

Two host profiles are illustrated here. The first one is named
\name{linux/ia32.ia32} while the second's name is \name{linux/ia32.mips64}.

For example, the \name{linux/ia32.mips64} \name{host} profile represents a
\name{Linux} operating system running on a \name{Intel 32-bit}
microprocessor while kaneton is built for a \name{MIPS 64-bit} target
architecture. This profile is likely to include the \location{linux.*} files
of the parent directory since there are not much difference between all the
\name{linux/*.*} \name{host} profiles. However, such a profile will
certainly redefine the binary paths of the C compiler, linker \etc{} in order
to produce \name{MIPS 64-bit} binary objects.

To conclude, the \name{host} sub-profile is accessed by the following
construct:

\begin{verbatim}
  profile/host/${KANETON_HOST}/${KANETON_ARCHITECTURE}
\end{verbatim}

With, for instance, the following values:

\begin{verbatim}
  KANETON_HOST = linux/ia32
  KANETON_ARCHTECTURE = ia32/educational
\end{verbatim}

Note that the possibility to include files in the configuration syntax allows
very similar profiles to share a huge amount of definitions.

% boot

\subsubsubsection{Boot}

The \name{boot} profile is both used for configuring the boot components
such as the \name{loader} \etc{} but also to set up the way a bootable
system is built.

% kaneton

\subsubsubsection{kaneton}

The \name{kaneton} profile is composed of four sub-profiles: \name{core},
\name{machine}, \name{library} and \name{modules} for the \name{kaneton}
microkernel sub-components respectively. Likewise, \name{machine} is
sub-divided into \name{platform}, \name{architecture} and \name{glue}.

The \name{core} sub-profile contains variables for parameterizing the
kaneton core. The \name{platform} and \name{architecture} sub-profiles
focus on the configuration of the platform- and architecture-dependent code
of the kaneton microkernel, respectively.

The user-defined shell variables \code{\$\{KANETON\_PLATFORM\}} and
\code{\$\{KANETON\_ARCHITECTURE\}} are used to address the \name{platform}
and \name{architecture} sub-profiles, respectively.

% user

\subsubsubsection{User}

Let us suppose that a developer would like the kaneton microkernel to
use a specific memory management entirely based on a \name{Slab Allocator}
and with all microprocessor optimisations enabled. These user-specific
configurations are actually allowed by the \name{user} profile.

The user-defined shell variable \code{\$\{KANETON\_USER\}} defines the name
of the \name{user} profile. This profile contains user-specific
configurations allowing a contributor to overwrite lower-level layer defintions
in order to specialise a behaviour.

The kaneton project also provides a tool allowing users to configure
their development environment. This tool is named \name{configure} and is
available from the kaneton project root directory. For more information about
this tool, please refer to \reference{Section \ref{section:configure}}.


%
% requirements
%

\subsubsection{Requirements}

The whole kaneton development environment needs exactly two fundamental tools
to work. The first one is \name{GNU make}, used to build powerful make files,
and the second one is \name{Python}, used to write portable scripts. If an
operating system has these two tools, then kaneton can certainly be developed
on it.

As said previously, the user has to specify some shell variables which are
used by the kaneton development environment engine. These variables are
described below:

\begin{itemize}
  \item
    \code{\$\{KANETON\_USER\}}: the name of the kaneton developer.

    A \name{user} profile name must be composed of the first name, a period
    and finally, the last name of the developer.
  \item
    \code{\$\{KANETON\_HOST\}}: the name of the host which is composed of
    a couple operating system/microprocessor.
  \item
    \code{\$\{KANETON\_PYTHON\}}: the path of the python binary.

    This path is required since the very first scripts which set up the
    configured environment are written in \name{Python}.
  \item
    \code{\$\{KANETON\_PLATFORM\}}: the name of the target platform.
  \item
    \code{\$\{KANETON\_ARCHITECTURE\}}: the name of the target architecture.
\end{itemize}

Note that once the configured environment is set up, these variables are
no longer used by the kaneton environment engine. Indeed, instead, the
kaneton environment operations are based on the \name{host} profile on
which rely the configured environment.

The profiles names must all be in lowercase letters. Below are some examples
of what could contain these variables:

\begin{verbatim}
  KANETON_USER='julien.quintard'

  KANETON_HOST='linux/ppc'
  KANETON_HOST='windows~cygwin/ia32'

  KANETON_PYTHON='/usr/bin/python'

  KANETON_PLATFORM='ibm-pc'
  KANETON_PLATFORM='sgi/o2'
  KANETON_PLATFORM='sgi/octane'

  KANETON_ARCHITECTURE='mips64'
  KANETON_ARCHITECTURE='ia32/educational'
  KANETON_ARCHITECTURE='ia32/smp'
\end{verbatim}

%
% organisation
%

\subsubsection{Organisation}

The development environment configuration files and scripts are located in
the \name{environment/} directory. The directory contains the three
following scripts:

\begin{verbatim}
  critical.py
  initialize.py
  clean.py
\end{verbatim}

The \location{critical.py} script essentially generates a configured
development environment. The result of this generation are two files called
\location{env.mk} and \location{env.py} which contains the configured
environment variables and functions for the \name{Make} files and \name{Python}
scripts, respectively. This file is called critical because it does not rely
on the portable development environment as it generates it.

The \location{initialize.py} script relies on the file \location{env.py}
previously generated. This script set up everything necessary for building the
kaneton microkernel based on the configured environment.

Finally, the \location{clean.py} script cleans everything installed by the
\location{initialize.py} script and removes the generated configured
environment files.

The generation of the configured environment is done by going through
the configuration files of all the profiles and sub-profiles associated
to the user configuration. In other words, the kaneton environment engine
processes the configuration files according to the layered organisation
described below, starting with the lowest-level layer thourgh the highest one.

\begin{verbatim}
  profile/
  profile/host
  profile/host/${KANETON_HOST}.${KANETON_ARCHITECTURE}
  profile/boot
  profile/boot/${KANETON_PLATFORM}.${KANETON_ARCHITECTURE}
  profile/kaneton
  profile/kaneton/core
  profile/kaneton/machine
  profile/kaneton/machine/platform
  profile/kaneton/machine/platform/${KANETON_PLATFORM}
  profile/kaneton/machine/architecture
  profile/kaneton/machine/architecture/${KANETON_ARCHITECTURE}
  profile/kaneton/machine/glue
  profile/kaneton/machine/glue/${KANETON_PLATFORM}.${KANETON_ARCHITECTURE}
  profile/kaneton/library
  profile/kaneton/modules
  profile/user
  profile/user/${KANETON_USER}         
\end{verbatim}

In this layered organisation, a variable defined in, for instance, the
\name{host} profile could be overwritten anywhere in the upper-level layers
\location{kaneton/},
\location{kaneton/architecture/\$\{KANETON\_ARCHITECTURE\}/}, \location{user/}
\etc{}

The \name{host} and \name{kaneton} profiles are theoretically completed
separated. However, the environment engine does not check for such
unauthorised overridings. Therefore the \name{core} configuration could
override a variable previously defined in the \name{host} profile.

Finally, the \name{user} profile can override any definition to adjust the
environment to its needs.

The environment engine looks for the following types of file in the
kaneton environment profile directories:

\begin{itemize}
  \item
    \location{.conf}: the \name{configuration} files contains variable
    definitions. These files are gathered by the development environment
    engine for generating the configured environment files.
  \item
    \location{.desc}: these \name{description} files contain descriptions of
    the variables of the current profile or sub-profile. These descriptions
    are used by the \name{configure} tool.
  \item
    \location{.mk}: the \name{Make} files usually contain the implementation
    of the kaneton \name{Make} interface.
  \item
    \location{.py}: the \name{Python} files usually contain the
    implementation of the kaneton \name{Python} interface.
\end{itemize}

The engine supposes that there is no variable or function overriding in
a single profile. More precisely, if there are more than a single
configuration file in a directory, the engine cannot guarantee anything
on the order these files will be processed. As a result, the overridings
could differ depending on the processing order.

Besides, although the environment engine gathers every \name{configuration}
files it finds in the environment profiles directory, it is highly recommended
to provide a single \name{configuration} file per profile directory. This
file should be named according to the name of its profile. For more details,
take a look at the environment directory which contains existing profiles.

Moreover, the \name{configure} tool requires the \name{user} profile to
contain a single \name{configuration} file named, as explained above,
\location{\$\{KANETON\_USER\}.conf}.

The kaneton development environment engine first gathers the
\name{configuration} files and process them creating an in-memory list of
configuration variables. Then it generates the configured environment files
\location{env.mk} and \location{env.py}. Indeed, the engine outputs the
configuration variables in each file and then append the content of the
\name{Make} files and \name{Python} files of the profiles to the
configured environment files \location{env.mk} and \location{env.py},
respectively.

Note that a special rule is included in \location{env.mk} so that the
configured files are regenerated if the environment engine detects that
an environment file has been changed since the last initialization.

%
% syntaxes
%

\subsubsection{Syntaxes}

% description

\subsubsubsection{Description}

The \name{description} files describe the environment variables in order
to specify what kind of value a variable can take \etc{}

The description syntax is based on the \name{YAML} language.

Examples of variable descriptions named \code{\_FOO\_}, \code{\_BAR\_}
and \code{\_CHICHE\_} are given next:

\begin{verbatim}
  #
  # _FOO_
  #
  - variable: _FOO_
    string: the foo flag
    type: set
    values:
      Off: -D_FOO_FLAG_=0
      On: -D_FOO_FLAG_=1
    description: |
      This is a description of a two-state variable _FOO_.

  #
  # _BAR_
  #
  - variable: _BAR_
    string: the bar parameter
    type: set
    values:
      Simple: ${BAR_SIMPLE}
      Normal: ${BAR_NORMAL}
      Optimised: ${BAR_OPTIMISED}
    description: |
      This is another parameter which can take three values: simple,
      normal and optimised.

  #
  # _CHICHE_
  #
  - variable: _CHICHE_
    string: the most powerful optimisation
    type: any
    description: This is the magic kaneton optimisation.
\end{verbatim}

Note that the environment engine never takes these descriptions into account.
Indeed, this is the role of the \name{configure} tool.

In this syntax, variables are classified according to the type of value
they can take: \code{set} and \code{any}.

A \code{set} variable can take any value in a given list of values. In
this case, the \code{values} field contains couples string/value the
variable can take. The \code{string} is displayed by the \name{configure}
tool while the \code{value} is assigned to the variable.

Finally, a \code{any} variable represents a variable which can take any
value.

The \code{string} fields were introduced to avoid displaying internal
non user-friendly variable names and/or values. Therefore, the
\name{configure} tool will always display literal \code{strings}
rather than variable names or values which are likely to do not make any
sense to the user.

% configuration

\subsubsubsection{Configuration}

The \name{configuration} files contains the actual variable definitions
through a very simple syntax.

The syntax allows both assignments and completion as shown in the next example:

\begin{verbatim}
  FOO = bar
  FOO += baz
  FOO = kaneton
\end{verbatim}

The \code{FOO} variable first takes the initial value \code{bar}. Then,
the value \code{baz} is added to the previous \code{FOO}'s value
leading the the value \code{bar baz}. Finally, the last assignment
overwrites the previous definition by setting the value of \code{FOO}'s
variable to \code{kaneton}.

The configuration syntax enables variable references. These references can
be both environment variable or shell variable. The following example
illustrates this.

\begin{verbatim}
  BAR = ${FOO} is a very powerful microkernel
  SH = the shell currently used is $(SHELL)
\end{verbatim}

The reader certainly notice the \code{\$\{\}} construct is used to reference
a kaneton environment variable while the \code{\$()} one references a shell
variable.

Finally, a configuration file can also include another file using the
\code{include} statement:

\begin{verbatim}
  include ../an/other/file/far/../far/../away
\end{verbatim}

This construct is very useful to centralize definitions common to multiple
profiles or sub-profiles in a single location.

Note that kaneton environment variables start and end with an underscore
to avoid naming collisions. Another solution would have been to use a
prefix \code{KANETON\_} as it is stipulated in the general kaneton rules
but this would have led to very long names.

% make

\subsubsubsection{Make}

The \name{Make} files must implement the whole kaneton \name{Make}
interface which will be described next.

The syntax used in these files is based on the \name{GNU Make} syntax.

% python

\subsubsubsection{Python}

The \name{Python} files must implement the whole kaneton \name{Python}
interface.

The syntax used in these files is based on the \name{Python} syntax.

%
% interfaces
%

\subsubsection{Interfaces}

% make

\subsubsubsection{Make}

In this section we will detail the make interface that every host profile
must implement. The reader should look closer to the host profiles already
implemented.

Since the \name{GNU Make} syntax does not provide any name space feature,
every kaneton \name{Make} function is prefixed by \code{env\_} in order
to avoid name conflicts.

Note that the \name{Make} development environment must take care of
setting the \code{PYTHONPATH} shell environment variable with a value
including the \code{\_PYTHON\_INCLUDE\_DIR\_} kaneton environment variable
so that scripts can use kaneton \name{Python} packages.

\function{}{env\_perform}{\argument{command}}
         {
           This function performs an action according to the given
           \argument{command} argument.

           \-

           Additionally, if the \code{\_OUTPUT\_} environment variable is
           set to \code{\$\{\_OUTPUT\_VERBOSE\_\}}, the function displays the
           command on the output before performing it.
         }

\function{}{env\_display}{\argument{color},
                          \argument{action},
                          \argument{file},
                          \argument{indentation},
                          \argument{options}}
         {
	   This function displays a message representing an action performed
	   by the kaneton \name{Make} interface.

	   \-

	   The option \code{\$(OPTION\_NO\_NEWLINE)} can be used not to
	   output the trailing newline.
	 }

\function{}{env\_cd}{\argument{directory},
                     \argument{options}}
         {
	   This function changes the current working directory.
	 }

\function{}{env\_pull}{\argument{file},
                       \argument{options}}
         {
	   This function returns the content of the \argument{file}.
	 }

\function{}{env\_launch}{\argument{file},
                         \argument{arguments},
                         \argument{options}}
         {
	   This function launches a new program/script/make \etc{}

	   \-

	   This function must look at the file name in order to determine
	   how to launch it. Moreover, the function must move the the
	   directory where is located the file before launching it.
	 }

\function{}{env\_preprocess}{\argument{preprocessed file},
                             \argument{c file},
                             \argument{options}}
         {
	   This function launches the C preprocessor the \argument{c file}
	   and generates the \argument{preprocessed file}.
	 }

\function{}{env\_compile-c}{\argument{object file},
                            \argument{c file},
                            \argument{options}}
         {
	   This function compiles a \argument{c file} generating an
	   \argument{object file}.
	 }

\function{}{env\_lex-l}{\argument{c file},
                        \argument{lex file},
                        \argument{options}}
         {
	   This function generates a \argument{c file} from a
	   \argument{lex file}.
	 }

\function{}{env\_assemble-S}{\argument{object file},
                             \argument{S file},
                             \argument{options}}
         {
	   This function assembles an \argument{S file}.
	 }

\function{}{env\_static-library}{\argument{static library file name},
                                 \argument{object files and/or libraries},
                                 \argument{options}}
         {
	   This function builds a static library from object files.
	 }

\function{}{env\_dynamic-library}{\argument{dynamic library file name},
                                  \argument{object files and/or libraries},
                                  \argument{options}}
         {
	   This function builds a dynamic library from \argument{object
           files and/or libraries}.
	 }

\function{}{env\_executable}{\argument{executable file name},
                             \argument{object files and/or libraries},
                             \argument{layout file},
                             \argument{options}}
         {
	   This function builds a executable from object files and/or
	   libraries. The \argument{layout file} describes where to
	   place the different data: code, read-only data, stack \etc{}

	   \-

	   The option \code{\$(ENV\_OPTION\_NO\_STANDARD)} tells the function
	   not to use the operating system standard stuff: libraries, includes
	   \etc{}
	 }

\function{}{env\_archive}{\argument{archive file name},
                          \argument{object files},
                          \argument{options}}
         {
	   This function builds an archive of objects from multiple
	   \argument{object files}.
	 }

\function{}{env\_remove}{\argument{files},
                         \argument{options}}
         {
	   This function removes the files in the list.
	 }

\function{}{env\_purge}{}
         {
	   This function just cleans the current working directory from
	   unecessary files.
	 }

\function{}{env\_prototypes}{\argument{files},
                             \argument{options}}
         {
	   This function generates prototypes in relation with the given
	   \argument{files}.
	 }

\function{}{env\_headers}{\argument{files},
                          \argument{options}}
         {
	   This function generates header dependencies for the \argument{files}
	   by building a \name{Make} dependency file named
	   \code{\$\{\_DEPENDENCY\_MK\_\}}.

	   \-

	   The generated files \code{\$\{\_DEPENDENCY\_MK\_\}} are removed by
	   the environment engine when cleaning the configured development
	   environment.
	 }

\function{}{env\_version}{\argument{file}}
         {
	   This function generates a version \argument{file} from the operating
	   system's informations: user, host, date \etc{}
	 }

\function{}{env\_link}{\argument{link},
                       \argument{file},
                       \argument{options}}
         {
	   This function creates a \argument{link} to the \argument{file}.
	 }

\function{}{env\_compile-tex}{\argument{file},
                              \argument{options}}
         {
	   This function compiles the file \argument{file}.tex and
	   generates a readable document.
	 }

\function{}{env\_document}{\argument{file},
                           \argument{options}}
         {
	   This function builds a document by calling the
           \code{env\_compile-tex()} function.

	   \-

           The option \code{\$(ENV\_OPTION\_PRIVATE)} configures the document
	   by setting the {\LaTeX} definition \code{$\backslash$mode} to
	   the value \code{private}. This option was introduced to deal with
	   documents which contain information which must be kept private to
	   the students.

	   \-

	   Note that a temporary file named \code{\$\{\_DEPENDENCY\_TEX\_\}}
	   is created by this functions storing the {\LaTeX} definition
	   \code{$\backslash$mode}. The developer should take care of
	   removing this file in the \code{clear} \name{Make} file rule.
	 }

\function{}{env\_view}{\argument{file},
                       \argument{options}}
         {
	   This function launches a viewer for the readable document
	   generated by the function \code{env\_compile-tex()}.

	   \-

	   Note that file, as for the \code{env\_compile-tex()} function,
	   does not have any filename suffix.
	 }

% python

\subsubsubsection{Python}

In this section we will detail the kaneton \name{Python} interface that
every \name{host} profile must implement.

The \name{Python} language was designed in a portable way. For this
reason, the major part of the \name{Python} interface is implemented
by the \name{host} generic profile.

Note that the \name{Python} language provides modularity through packages.
Therefore, each \name{Python} script has to import the \code{env} package
generated by the development environment engine: \location{environment/env.py}.
Then, environment functions and variables are accessed through this package.

Below are described the functions implemented by the \code{env} package.

Note that the \name{Python} development environment must take care of
setting the \code{PYTHONPATH} shell environment variable with a value
including the \code{\_PYTHON\_INCLUDE\_DIR\_} kaneton environment variable
so that scripts can use kaneton \name{Python} packages.

\function{}{display}{\argument{header},
                     \argument{text},
                     \argument{options}}
         {
	   This function outputs some text to the screen depending on the
	   \argument{header}: \code{HEADER\_NONE}, \code{HEADER\_OK},
	   \code{HEADER\_ERROR} or \code{HEADER\_INTERACTIVE}.
	 }

\function{}{pull}{\argument{file},
                  \argument{options}}
         {
	   This function returns the content of the \argument{file}.
	 }

\function{}{push}{\argument{file},
                  \argument{content},
                  \argument{options}}
         {
	   This function writes the \argument{content} in the \argument{file}.
	 }

\function{}{temporary}{\argument{options}}
         {
	   This function creates a temporary file system object.

	   \-

	   The options \code{OPTION\_FILE} and \code{OPTION\_DIRECTORY}
	   specify which type of object to create.
	 }

\function{}{cwd}{\argument{options}}
         {
	   This function returns the path of the current working directory.
	 }

\function{}{input}{\argument{options}}
         {
	   This function waits for an input from the user.
	 }

\function{}{launch}{\argument{file},
                    \argument{arguments},
                    \argument{options}}
         {
	   This function launches a new program/script/make file \etc{}

	   \-

	   This function must look at the file name in order to determine
	   how to launch it. Moreover, the function must move the the
	   directory where is located the file before launching it.

           \-

           The option \code{OPTION\_QUIET} makes the \code{launch()} function
           do not print anything on the output screen.
	 }

\function{}{copy}{\argument{source},
                  \argument{destination},
                  \argument{options}}
         {
	   This function copies the file \argument{source} to
	   \argument{destination}.
	 }

\function{}{link}{\argument{source},
                  \argument{destination},
                  \argument{options}}
         {
	   This function builds a link between the file \argument{source}
	   and the file \argument{destination}.
	 }

\function{}{remove}{\argument{target},
                    \argument{options}}
         {
	   This function removes the \argument{target} which can be either
	   a file or a directory.
	 }

\function{}{list}{\argument{directory},
                  \argument{options}}
         {
	   This function lists the file system objects contains in the
	   \argument{directory}.

	   \-

	   The options \code{OPTION\_FILE} and \code{OPTION\_DIRECTORY}
	   specify which type of object to list.
	 }

\function{}{cd}{\argument{directory},
                \argument{options}}
         {
	   This function changes the current working directory to
	   \argument{directory}.
	 }

\function{}{search}{\argument{directory},
                    \argument{pattern},
                    \argument{options}}
         {
	   This function looks for files matching the given \argument{pattern}.

	   \-

	   The options \code{OPTION\_FILE} and \code{OPTION\_DIRECTORY}
	   specify which type of object to list while the
	   \code{OPTION\_RECURSIVE} option tells the function to explore
	   the whole file system sub-tree.
	 }

\function{}{pack}{\argument{directory},
                  \argument{file},
                  \argument{options}}
         {
	   This function makes an archive \argument{file} of the
	   \argument{directory}.
	 }

\function{}{unpack}{\argument{file},
                    \argument{directory},
                    \argument{options}}
         {
	   This function extracts the archive \argument{file} into the
	   \argument{directory}, if specified.
	 }

\function{}{mkdir}{\argument{directory},
                   \argument{options}}
         {
	   This function builds a new directory named \argument{directory}.
	 }

\function{}{load}{\argument{file},
                  \argument{device},
                  \argument{path},
                  \argument{options}}
         {
	   This function copies the \argument{file} on the specificed
	   \argument{device}, more precisely at the location \argument{path}.
	   The device can be virtual: an image.

	   \-

	   The options \code{OPTION\_DEVICE} and \code{OPTION\_IMAGE}
	   specify on which type of device the file must be copied.
	 }

\function{}{stamp}{\argument{options}}
         {
	   This function returns a current date.
	 }

\function{}{record}{\argument{transcript},
                    \argument{options}}
         {
	   This function starts recording a session and transcripts it
           into \argument{transcript}.
	 }

\function{}{play}{\argument{transcript},
                  \argument{options}}
         {
	   This function plays a previously recorded \argument{transcript}.
	 }

\function{}{locate}{\argument{file},
                    \argument{options}}
         {
	   This function tries to locate the program \argument{file}
	   on the system.
	 }

\function{}{path}{\argument{path},
                  \argument{options}}
         {
	   This function returns information on the given \argument{path}.

	   \-

	   The options \code{OPTION\_FILE} and \code{OPTION\_DIRECTORY}
	   specify which information the caller is interested in. The option
           \code{OPTION\_EXIST} indicates whether the \argument{path} object
           exists or not.
	 }

\function{}{info}{\argument{options}}
         {
	   This function returns information on the system.

	   \-

	   The option \code{OPTION\_CURRENT\_DIRECTORY} returns the
           sequence of characters used for accessing the current directory.
	 }
