%
% ---------- header -----------------------------------------------------------
%
% project       kaneton
%
% license       kaneton
%
% file          /home/mycure/kaneton/view/book/development/mailing-list.tex
%
% created       julien quintard   [thu may 24 19:55:18 2007]
% updated       julien quintard   [sat oct 20 14:24:08 2007]
%

%
% ---------- mailing-list -----------------------------------------------------
%

\subsection{Mailing-List}
\label{section:mailing-list}

Because people do not want to use several communication tools: email,
newsgroup, forum etc.. and because everybody has an email address, the
kaneton people communicate through a mailing-list.

This mailing-list is actually a \textit{Google Group}. The kaneton
project relies on four distinct communication groups:

\begin{itemize}
  \item
    \textbf{kaneton-announcements} which is used to communicate with all the
    people involved in the kaneton projects.

    \-

    This group is obviously private to the people involved.
  \item
    \textbf{kaneton-developers} is dedicated to the communication between the
    people involved in the development of the kaneton microkernel research
    project.

    \-

    This group is therefore private.
  \item
    \textbf{kaneton-teachers} is reserved to people involved in teaching
    with the kaneton educational project.

    \-

    This group is also private.
  \item
    \textbf{kaneton-students} can be used in an absolutely free-way by
    students for communicating about their kaneton educational implementation.

    \-

    Anybody can join this group.
\end{itemize}

In addition, another \textit{Google Group} is used to communicate with people
involved in the sibling projects: \textit{k} and \textit{kayou}:

\begin{itemize}
  \item
    \textbf{k-kaneton-kayou} must therefore be used to make announcements
    and/or talk about general subjects.

    \-

    This group is obviously private to the developers, teachers etc. involved
    in one or more projects including \textit{k}, \textit{kaneton} and/or
    \textit{kayou}.
\end{itemize}

Needless to say, community rules discussed in \textit{Chapter
\ref{chapter:community}} must be followed when communicating on the kaneton
mailing-lists. Every contributor is welcomed to give its point of view, to
ask questions etc.. but this must be done with politness, respect and humility.

Everyone communicating through the mailing-list must read the
\textit{Netiquette} which describes the rules inherent to the communication
on the Internet. Especially, people should take care of writing messages in
respect of the \textit{80} columns; and should always cut off useless parts
in previous messages when responding.

It is likely that a real-time communication tool will be very useful in the
future, an \textit{IRC} channel for instance. However, communicating on
these extra media will not be mandatory unless kaneton people decide so.

kaneton people are asked to use the mailing-list communication medium
in a perfect way as it is the unique intra-communication channel. In
addition then, contributors must read their emails on a regular-basis
as some people rely on the decision of others.

The \textit{kaneton-students} mailing-list must be used carefully as well. As
an example, people should never paste pieces of source code or ask
questions implying an answer with the solution. Even if it is a free
group, people abusing of this communication channel could be easily banned.

People must always respond in the appropriate discussion. If, in a discussion,
a different subject is discussed, then, one of the contributor must create
a new discussion in order facilitate the communication.

Finally, the discussion subjects must be tagged like the following examples:

\begin{verbatim}
  [ia32/optimised] mapping issues

  [segment] segment_clone() :: bug?

  [research] new paper about OS design
\end{verbatim}

There is no list of official tags. The users are simply asked to make
their discussion subjects as clear as possible to simplify the task
consisting in looking for old topics into the archives. Indeed, remember
that newcomers should - if they respect the rules - look at the archives
to avoid discussing an old subject on the mailing-list.
